---
title: "Exploratory Data Analysis"
author: "Peter Chong"
output: github_document
---

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(ggridges)
```

```{r}
#Input files
df <- read.csv("/Users/wengliangchong/Desktop/RunTheData/MarathonData.csv", header = T, stringsAsFactors = F)
```

# The Dataset

We have a dataset of 75652 runners information across 4 years (2015~2018). Let's look at the head of data.

```{r}
head(df)
```

Let's know the runners a little better

### Proportion of Gender
```{r}
df1 <- df %>%
  group_by(Gender) %>%
  count() %>%
  ungroup() %>%
  mutate(per=`n`/sum(`n`)) %>% 
  arrange(desc(Gender))
df1$label <- scales::percent(df1$per)
ggplot(data=df1) + 
  geom_bar(aes(x="", y=per, fill=Gender), stat="identity", width = 1, alpha=0.7) +
  coord_polar("y", start=0) +
  geom_text(aes(x=1, y = cumsum(per) - per/2, label=label)) + 
  theme(axis.title.x = element_blank(), axis.title.y = element_blank(),
        panel.grid = element_blank(), axis.text.x = element_blank(),
        axis.ticks = element_blank(), panel.border = element_blank(), axis.text.y = element_blank())
```

This graph is very simple, it show us that male has a higher proportion across four years. Let's see if that is the same case for each year.

### Proportion of Gender according to year
```{r}
df2 <- df %>%
  group_by(Year, Gender) %>%
  count() %>%
  mutate(count=n) %>%
  ungroup()
ggplot(df2, aes(x=Year, y=count, fill=Gender)) + 
  geom_bar(stat="identity", position=position_dodge())
```

Yes, it is. Male participants are higher than female participants every year. Let's see if male participants finish the marathon faster than female participants.

### Finish time distribution by gender
```{r}
ggplot(df, aes(x=as.POSIXct(Net.Time,format="%H:%M:%S"), fill = Gender, color=Gender)) + 
  geom_histogram(aes(y=..density..),alpha=0.1, position="identity", binwidth = 250) +
  geom_density(alpha=0.6, fill=NA) + xlab("Net Finish Time") + ylab("") +
  theme(legend.position = c(0.9, 0.8), axis.text.y = element_blank(), axis.ticks = element_blank())
```

Yes, male participants generally finish faster than female participants. Interestingly, the finish time follows a normal distribution for both men and women. Next, let's see the age distribution across the four years.

### Age Distribution
```{r}
ggplot(df, aes(x=Age, fill=Gender, color=Gender)) +
  geom_histogram(aes(y=..density..),alpha=0.1, position = "identity", binwidth = 1) +
  geom_density(alpha=0.6, fill=NA) + ylab("Distribution") +
  theme(legend.position = c(0.9, 0.8)) + scale_x_continuous(breaks=round(seq(0, 90, by = 10),1))
```

We can observe that there is a high interest in marathon for the people ages 15~18 but significantly reduced for the people ages 19~25.  
This age distribution is very interesting as neither men or women curves belong to a normal distribution. Although I do not know the reason behind the spike in interest for the people ages 15~18, I would assume people ages 19~25 are busy with university. Let's see if this is the same across four years.

### Age Distribution by Year
```{r}
df$Year <- as.character(df$Year)
ggplot(df, aes(x = Age, y = Year, fill=stat(y))) + 
  geom_density_ridges(stat="binline", binwidth = 1, scale=1.5, alpha=0.3) +
  scale_fill_viridis_c(option = "C") + theme(legend.position = "none")
df$Year <- as.numeric(df$Year)
```

Yes, it is true that the age distribution is very similar across all four years. Unfortunately, I have no idea the reason behind this trend. Next, we map the finish time to the Age distribution.

### Net Finish Time against Age
```{r}
ggplot(df, mapping = aes(x = Age, y = as.POSIXct(Net.Time,format="%H:%M:%S"))) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon") + ylab("Net Finish Time") +
  scale_fill_continuous(low="lightblue", high="darkblue") + 
  scale_x_continuous(breaks=round(seq(10, 70, by = 10),1)) + theme(legend.position = "none")
```

There are two different "typical runner" scenarios when we're looking at the finish times verses age. The early 30s that run in under 5 hours and the teenagers that runs in under 7 hours.  
  
Let's plot age versus net finish times in a different way to see if we can learn something new.

### Ridge plot: Age Group against Finish Time
```{r}
ggplot(df, aes(x = as.POSIXct(Net.Time,format="%H:%M:%S"), y = AgeGroup, fill=stat(y))) + 
  geom_density_ridges(stat="binline", binwidth = 100, scale=1.5, alpha=0.3) +
  scale_fill_viridis_c(option = "C") + theme(legend.position = "none") + 
  ylab("Age Group") + xlab("Net Finish Time")
```

We notice that for age group below 24 and above 65, we can see a shift in finish times to the right.  
We can also observe that the mean finish time excluding the age group 1 ~ 24 and 65 ~ 90 is around 5 hours. One thing that I would like to look into is the psychology around the five hour mark.  
  
Let’s look closer at the finish times around 5 hours.

### Histogram at time 4:55:00 to 5:05:00
```{r, warning=FALSE}
colors <- c(rep("lightgreen",21), rep("lightblue",20))
ggplot(df, aes(x=as.POSIXct(Net.Time,format="%H:%M:%S"))) + 
  geom_histogram(binwidth = 15, color="black", fill=colors) + xlab("Net Finish Time") + 
  xlim(c(as.POSIXct("04:55:00",format="%H:%M:%S"), as.POSIXct("05:05:00",format="%H:%M:%S")))
```

As a runner, we would always try to run within a specific hours of time (eg. within 3 hours or 4 hours). Let's see if it is true that the hourly mark will motivate people to run faster and cross the finishing line before hitting the mark. Let's count the proportion of people finishing between 4:58:00 to 5:00:00 and 5:00:00 to 5:02:00.

```{r}
sub5 <- nrow(df[which(df$X42ks > 17880 & df$X42ks <= 18000),])
notsub5 <- nrow(df[which(df$X42ks > 18000 & df$X42ks <= 18120),])
round(sub5/(sub5+notsub5)*100,1)
```

So for people that cross the finishing line in between 4:58:00 to 5:02:00, 53% of them crossed the finishing line before 5 hour mark.  
  
Next, I want to know if we keep changing our pace for every leg, does it affect our finishing time?

```{r}
mean <- c()
sd <- c()
for (i in 1:nrow(df)) {
  speed <- c(df$first5kpace[i], df$second5kpace[i], df$third5kpace[i], df$fourth5kpace[i],
             df$fifth5kpacec[i], df$sixth5kpace[i], df$seventh5kpace[i], df$eighth5kpace[i], df$last2kpace[i])
  mean[i] <- mean(speed)
  sd[i] <- sd(speed)
}
df$mean <- mean
df$sd <- sd
```

### Pace change
```{r}
ggplot(df, aes(Age, as.POSIXct(Net.Time,format="%H:%M:%S"))) +
  geom_point(aes(color = sd), alpha=0.7) + scale_color_gradient(low="#FDFBBB", high="#000000") +
  ylab("Net Finish Time") + scale_x_continuous(breaks=seq(5,90,10))
```

We plot the standard deviation of the pace for every runner. The darker the circle means that the more uneven the pace the runner ran. We can see that the more uneven the runner ran, the slower they finish the marathon. Hence, it means that it would be advantageous if we keep an even pace.  
  
Now, let's look at how runners vary their pace, and if there are any difference between men and women.
```{r}
d5k <- d10k <- d15k <- d20k <- d25k <- c()
d30k <-d35k <- d40k <- d42k <- c()
for (i in 1:nrow(df)) {
  d5k[i] <- (df$first5kpace[i] - df$mean[i])/df$mean[i]*100
  d10k[i] <- (df$second5kpace[i] - df$mean[i])/df$mean[i]*100
  d15k[i] <- (df$third5kpace[i] - df$mean[i])/df$mean[i]*100
  d20k[i] <- (df$fourth5kpace[i] - df$mean[i])/df$mean[i]*100
  d25k[i] <- (df$fifth5kpace[i] - df$mean[i])/df$mean[i]*100
  d30k[i] <- (df$sixth5kpace[i] - df$mean[i])/df$mean[i]*100
  d35k[i] <- (df$seventh5kpace[i] - df$mean[i])/df$mean[i]*100
  d40k[i] <- (df$eighth5kpace[i] - df$mean[i])/df$mean[i]*100
  d42k[i] <- (df$last2kpace[i] - df$mean[i])/df$mean[i]*100
}
```

```{r}
Checkpoint <- c("5 km", "10 km", "15 km", "20 km", "25 km", "30 km", "35 km", "40 km", "42 km")
Gender <- rep(c("Male", "Female"),each=9)
Diff <- c()
for (j in c(1,10)) {
  Diff[j] <- mean(d5k[which(df$Gender == Gender[j])])
  Diff[j+1] <- mean(d10k[which(df$Gender == Gender[j+1])])
  Diff[j+2] <- mean(d15k[which(df$Gender == Gender[j+2])])
  Diff[j+3] <- mean(d20k[which(df$Gender == Gender[j+3])])
  Diff[j+4] <- mean(d25k[which(df$Gender == Gender[j+4])])
  Diff[j+5] <- mean(d30k[which(df$Gender == Gender[j+5])])
  Diff[j+6] <- mean(d35k[which(df$Gender == Gender[j+6])])
  Diff[j+7] <- mean(d40k[which(df$Gender == Gender[j+7])])
  Diff[j+8] <- mean(d42k[which(df$Gender == Gender[j+8])])
}
df3 <- data.frame("Checkpoint" = rep(Checkpoint,2), "Diff" = Diff, "Gender" = Gender)
```

### Average pace profile by gender
```{r}
ggplot(data=df3, aes(x=Checkpoint, y=Diff, fill = Gender, color=Gender)) +
  geom_bar(stat="identity",position=position_dodge(), alpha=0.5) + 
  scale_x_discrete(limits=Checkpoint) + theme(legend.position = c(0.15, 0.8)) +
  ylab("Difference to average pace (%)") + scale_y_continuous(breaks=seq(-15,15,5))
```

Let's assume that every runner have a desired pace they want to run throughout the marathon and they would love to keep a consistent pace (also since we know having a consistent pace is advantageous). We assume that the desired pace is ran between km5 to km10. Now, let's see how much the pace change throughout the course compared to the desired pace.

### Percentage pace difference relative to 5k~10k pace by gender
```{r}
Leg <- c("km 10-15", "km 15-20", "km 20-25", "km 25-30", "km 30-35", "km 35-40", "km 40-42")
Gender <- rep(c("Male", "Female"),each=7)
Diff <- c()
for (j in c(1,8)) {
  Diff[j] <- mean(df$Hubris15[which(df$Gender == Gender[j])])
  Diff[j+1] <- mean(df$Hubris20[which(df$Gender == Gender[j+1])])
  Diff[j+2] <- mean(df$Hubris25[which(df$Gender == Gender[j+2])])
  Diff[j+3] <- mean(df$Hubris30[which(df$Gender == Gender[j+3])])
  Diff[j+4] <- mean(df$Hubris35[which(df$Gender == Gender[j+4])])
  Diff[j+5] <- mean(df$Hubris40[which(df$Gender == Gender[j+5])])
  Diff[j+6] <- mean(df$Hubris42[which(df$Gender == Gender[j+6])])
}
df4 <- data.frame("Leg" = rep(Leg,2), "Diff" = Diff, "Gender" = Gender)

ggplot(data=df4, aes(x=Leg, y=Diff, fill=Gender, color=Gender)) + ylab("Difference to 5k~10k pace (%)") +
  geom_bar(stat="identity",position=position_dodge(), alpha=0.5) + theme(legend.position = c(0.15, 0.8)) +
  scale_x_discrete(limits=Leg)
```

This plot shows the percentage difference between pace for a specific leg and the pace for the second leg. For example, between kilometer 20–25, men and women held around 10% slower pace compared to the second 5k, on average. It shows that men seem to burn out more than women do starting from km 20. This plot also shows that men try harder to keep their opening pace but suffer a larger pace reduction in the later stages of the race. Last but not least, women recovered and ran a similar pace as their opening pace at the last 2kms.  
  
Let's see if we have the same effect on age groups.

### Percentage pace difference relative to 5k~10k pace by age group
```{r}
AgeGroup <- rep(c("1 ~ 24", "25 ~ 34", "35 ~ 44", "45 ~ 54", "55 ~ 64", "65 ~ 90"), each=7)
Diff <- c()
for (j in c(1,8,15,22,29,36)) {
  Diff[j] <- mean(df$Hubris15[which(df$AgeGroup == AgeGroup[j])])
  Diff[j+1] <- mean(df$Hubris20[which(df$AgeGroup == AgeGroup[j+1])])
  Diff[j+2] <- mean(df$Hubris25[which(df$AgeGroup == AgeGroup[j+2])])
  Diff[j+3] <- mean(df$Hubris30[which(df$AgeGroup == AgeGroup[j+3])])
  Diff[j+4] <- mean(df$Hubris35[which(df$AgeGroup == AgeGroup[j+4])])
  Diff[j+5] <- mean(df$Hubris40[which(df$AgeGroup == AgeGroup[j+5])])
  Diff[j+6] <- mean(df$Hubris42[which(df$AgeGroup == AgeGroup[j+6])])
}
df5 <- data.frame("Leg" = rep(Leg,6), "Diff" = Diff, "AgeGroup" = AgeGroup)

ggplot(data=df5, aes(x=Leg, y=Diff, fill=AgeGroup, color=AgeGroup)) + ylab("Difference to 5k pace (%)") +
  geom_bar(stat="identity",position=position_dodge(), alpha=0.5) + theme(legend.position = c(0.15, 0.7)) +
  scale_x_discrete(limits=Leg)
```

A similar effect can be observed here. However, we notice that age group 1 ~ 24 burn out way more than any other group.  
  
Now, after having some EDA, we understood our data more now. Let's move on to model prediction to predict the finishing times.

